A Practical Optional Type System for Clojure

### 2012-11-01 dissertation/experiments.tex
Wording

As Typed Clojure is intended to be useful for practical purposes, it was important
to port existing code that utilized Java interoperability.
My porting effort attempted to follow how a real programmer might port code to Typed Clojure
and I describe my approach for each situation.

I ported a function from clojure.contrib.reflect\footnote{https://github.com/richhickey/clojure-contrib},
a Clojure library relying heavily on Java interoperability.
I chose \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, it exposes a wart of Clojure with \emph{Named} things,
and \lstinline|null| is a valid value in one place.

Before showing the implementation, there is a brief explanation of the relevant syntax.
Java methods are called using the \emph{dot} operator. If \lstinline|o| is an object \lstinline|(. o m a*)|
calls its method named \lstinline|m| passing \lstinline|a*| arguments. The method can be named first
using the equivalent sugar \lstinline|(.m o a*)|. Also, \lstinline|doto| is convenient 
notation for muliple method calls on the same object, presumably for side effects, and returns the original
object. For example, \lstinline|(doto o (.m1 a*) (m2 a*))| calls methods \lstinline|m1| and \lstinline|m2|
on \lstinline|o| and returns \lstinline|o|.

@@

\begin{itemize}
\item \lstinline|(into-array Class params)| becomes \lstinline|(into-array> Class Class params)|
      which is of type \lstinline|(Array Class Class)|.
\item \lstinline|(into-array Object args)| becomes \lstinline|(into-array> Object (U nil Object) args)|
      which is of type \lstinline|(Array Object (U nil Object))|.
\end{itemize}

>>>>

It is idiomatic and common in Clojure to interface with existing Java code
via Java interoperability.
Typed Clojure is intended to be useful for practical purposes, so it is important
to understand this common feature.
To test Typed Clojure's approach, 
existing code utilising Java interoperability was ported.
My porting effort attempted to follow how a real programmer might port code to Typed Clojure;
I describe this process step-by-step.

This section describes porting a function from \emph{clojure.contrib.reflect}
\footnote{\emph{clojure.contrib} Github project: https://github.com/richhickey/clojure-contrib},
a Clojure library that relies heavily on Java interoperability.
I chose to port \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, and \lstinline|null| is a valid value in one place.

Before showing the implementation, there is a brief explanation of the relevant syntax.
Java methods are called using the \emph{dot} operator. If \lstinline|o| is an object,
then \lstinline[mathescape]|(. o m $\overrightarrow{a}$)|
calls its method named \lstinline|m|, passing \lstinline[mathescape]|$\overrightarrow{a}$| as arguments. 
Syntactic sugar allows the method to be named first:
\lstinline[mathescape]|(.m o $\overrightarrow{a}$)|
is equivalent to
\lstinline[mathescape]|(. o m $\overrightarrow{a}$)|.
Also, \lstinline|doto| is convenient 
notation for chaining multiple method calls to the same object, presumably for side effects, and returns the original
object. For example, \lstinline[mathescape]|(doto o (.m1 $\overrightarrow{a}$) (.m2 $\overrightarrow{b}$))| 
calls methods \lstinline|m1| (passing \lstinline[mathescape]|$\overrightarrow{a}$| as arguments)
and \lstinline|m2| (passing \lstinline[mathescape]|$\overrightarrow{b}$| as arguments)
on \lstinline|o| in order, and returns \lstinline|o| with any side effects applied to it.

@@

In particular we change two invocations of \lstinline|into-array|.

\begin{itemize}
\item \lstinline|(into-array Class params)|, which creates an array of type \lstinline|Class[]| and populates it
with the elements of \lstinline|params|,
 becomes \lstinline|(into-array> Class Class params)|,
      which is of type \lstinline|(Array Class Class)|.
\item \lstinline|(into-array Object args)|, which creates an array of type \lstinline|Object[]|
and populates it with the elements of \lstinline|args|,
becomes \lstinline|(into-array> Object (U nil Object) args)|
      which is of type \lstinline|(Array Object (U nil Object))|.
\end{itemize}
